### 2025-08-13 paper/camera/sects/discussion.tex
prepare for camera ready

This scaling advantage is not observed in the Dispersion scenario, where complexity increases quadratically with agent count. Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (quantitative analysis with up to 20 agents is provided in \Cref{apx:experiments}). The primary computational constraint lies in the transformer architecture employed for the action world model, where sequence length scales linearly with the number of agents (specifically, sequence length = $n \times s$, where $n$ is the number of agents and $s$ is the number of training steps). With agent populations exceeding approximately 50 agents, the self-attention mechanism's $O(n^2)$ complexity leads to prohibitive memory requirements. For such large-scale multi-agent systems, linear attention mechanisms (e.g.,~\cite{Beltagy20}) represent a viable architectural alternative.

>>>>

This scaling advantage is not observed in the Dispersion scenario, where complexity increases quadratically with agent count. Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (see \Cref{apx:experiments} for quantitative analysis with up to 20 agents in appendix~\cite{appendix}). The primary computational constraint lies in the transformer architecture employed for the action world model, where sequence length scales linearly with the number of agents (specifically, sequence length = $n \times s$, where $n$ is the number of agents and $s$ is the number of training steps). With agent populations exceeding approximately 50 agents, the self-attention mechanism's $O(n^2)$ complexity leads to prohibitive memory requirements. For such large-scale multi-agent systems, linear attention mechanisms (e.g.,~\cite{Beltagy20}) represent a viable architectural alternative.

### 2025-08-13 paper/camera/sects/experiments.tex
prepare for camera ready

\paragraph{Scenarios}
We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
For more details about the choice of environments, see \Cref{apx:scenarios}.
We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details). These scenarios can be categorized along two dimensions:
\begin{compactitem}
    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state. In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
\end{compactitem}

@@

While the MLP architecture is the simplest, the Transformer baseline benefits from positional invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture. For more details on the architectures, see \Cref{apx:arch}.

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters. Hyperparameter choices are the result of a small grid search on the transport scenario.

@@

In this section, we compare the performance of \fname{}, PPO/MAPPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.

\paragraph{Dispersion}
In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable). In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation. In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior. In the third column (5 agents), PPO fails to converge, while \fname{} achieves high rewards (with no early stopping) but shows high variance due to increased environment complexity.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.75\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO/MAPPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
    \label{fig:experiments}
    \vspace{0.1cm}
\end{figure*}

>>>>

\paragraph{Scenarios}
We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
For more details about the choice of environments, see \Cref{apx:scenarios} in appendix~\cite{appendix}.
We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} in appendix~\cite{appendix} for details). These scenarios can be categorized along two dimensions:
\begin{compactitem}
    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state. In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
\end{compactitem}

@@

While the MLP architecture is the simplest, the Transformer baseline benefits from positional invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture. For more details on the architectures, see \Cref{apx:arch} in appendix~\cite{appendix}.

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds. See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} in appendix~\cite{appendix} for a complete list of hyperparameters. Hyperparameter choices are the result of a small grid search on the transport scenario.

@@

In this section, we compare the performance of \fname{}, PPO/MAPPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of achieved rewards over 3 runs. An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}.

\paragraph{Dispersion}
In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable). In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation. In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior. In the third column (5 agents), PPO fails to converge, while \fname{} achieves high rewards (with no early stopping) but shows high variance due to increased environment complexity.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.75\textwidth]{figs/main-transformer.pdf}
    \caption{Performance comparison of \fname{}, PPO/MAPPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings. The results show the mean and standard deviation of rewards over 3 runs. For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}.}%
    \label{fig:experiments}
    \vspace{0.1cm}
\end{figure*}
